Numerical simulation of the Cordilleran ice sheet \\ through the last glacial cycle

### 2014-07-08 kappa.tex
Develop glacial history.

% ----------------------------------------------------------------------
\subsection{Palaeo-glaciology}
% ----------------------------------------------------------------------

During the last glacial maximum, the Cordilleran ice sheet covered most of the
Canadian Cordillera, and some of the adjacent land in north-western contiguous
USA and Alaska (\cref{fig:map-cordillera}). The southern margin is the most
extensively studied. It consists of different piedmont lobes that have formed
at the outlet of the main valleys. The Puget lobe was the westernmost and
largest lobe. It covered the Puget Lowland, a region that currently includes
the cities of Vancouver and Seattle. East of the Cascade Range, were found the
Okanagan and Purcell lobes. The latter once dammed the Clark Fork River to form
Glacial Lake Missoula \citep{Pardee.1910}. As the ice sheet retreated, the lake
repeatedly and catastrophically drained to the Pacific Ocean
\citep{Bretz.1923,Waitt.1980}. These floods
contributed to shape a distinctive landscape known as the Channeled Scabland.

@@

Although the last glacial maximum extent of the Cordilleran ice sheet, apart of
its eastern margin, is relatively well known, the history of deglaciation is
heavily debated, while little information exists on earlier dynamics of ice
sheet.

>>>>

% ----------------------------------------------------------------------
\subsection{Glacial history}
% ----------------------------------------------------------------------

Earliest glaciofluvial deposits associated with the presence of an ice sheet in
the Cordilleran mountains were dated at $2.64^{+0.20}/_{-0.18}$\,Ma
\citep{Hidy.etal.2013}. This suggests that the North American Cordillera has
been glaciated many times, with the first glaciations occuring even before the
onset of the 100\,ka glacial cycle (\cref{fig:paleo-timeseries}). However, only
the last few of these episodes can be reconstructed from the geologic record,
whereas much of the older evidence has been overriden.

In the northern sector of the ice sheet, at least four glaciations have been
identified \mref. The two older episodes, often referred to as Reid and
pre-Reid, are older than the last glacial cycle \mref, and thus beyond the
scope of this study. The two younger glaciations, though, occurred both during
the last glacial cycle \mref. Indeed, the Cordilleran ice sheet did not keep
its maximum size during the entire glacial cycle. Instead, it is thought that,
for most of the last glacial cycle, ice cover was restricted to some of the
major mountain ranges of the Cordillera, while the maximum, ice-sheet stage,
corresponded to one or more short-lived expansion cycles.

The northern margin of the ice sheet suggests two such glacial episodes. The
first expansion cycle, the Gladstone glaciation, occurred at about 55-60\,ka,
during a local maximum of global ice volume on Earth known as the Marine
Isotope Stage (MIS)~4 \mref. However, evidence for the Gladstone glaciation has
not been found the southern part of the ice-covered region, making it difficult
to seize its magnitude. The second expansion cycle, the McConnell glaciation,
was dated at 18\,ka, which corresponds to the last global ice volume maximum
(MIS~2). In opposite to the Gladstone glaciation, the McConnell glaciation has
been correlated to a contemporaneous glacier advance of the southern margin,
known as the Fraser glaciation, and dated at 17\,ka. Although a slight
difference in timing may exists between the timing of maximum extent of the
northern and southern ice margins, the The McConnell and Fraser glaciation
correspond to the same expansion cycle, which resulted in the Last Glacial
Maximum (LGM) extent of the Cordilleran ice sheet. This is the period where
most geological observations point at.

It is generally thought that the Cordilleran ice sheet built from the
coalescence of several mountain-centred ice caps \citep{Davis.Mathews.1944}.
During the last glacial maximum, the Cordilleran ice sheet covered most of the
Canadian Cordillera, and some of the adjacent land in north-western contiguous
USA and Alaska (\cref{fig:map-cordillera}). The southern margin is the most
extensively studied. It consists of different piedmont lobes that have formed
at the outlet of the main valleys. The Puget lobe was the westernmost and
largest lobe. It covered the Puget Lowland, a region that currently includes
the cities of Vancouver and Seattle. East of the Cascade Range, were found the
Okanagan and Purcell lobes. The latter once dammed the Clark Fork River to form
Glacial Lake Missoula \citep{Pardee.1910}. As the ice sheet retreated, the lake
repeatedly and catastrophically drained to the Pacific Ocean
\citep{Bretz.1923,Waitt.1980}. These floods
contributed to shape a distinctive landscape known as the Channeled Scabland.

@@

The ice sheet is believed to have retreated quickly. Most of the evidence comes
from the Fraser Plateau, where it indicates that some of the highest relief
emerged from the ice prior to the lowland, possibly resulting in thin, stagnant
ice remnants in the interior
    \citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}.
Much less is known about ice sheet decay in the northern parts of the
ice-covered areas, but glacial landforms from the Liard Lowland suggest a
different mode of deglaciation, with active retreat of the ice lobe towards the
surrounding mountain ranges \citep{Margold.etal.2013}. Patterns of
deglaciation, as well as the possibility for a late-glacial re-advance, are
discussed in further detail with consideration of modelling results in \CCYC.
